---
title: "Introduction to PulmoDataSets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to PulmoDataSets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PulmoDataSets)
library(dplyr)
library(ggplot2)
```

# Introduction

The **PulmoDataSets** package offers a thematically rich and diverse collection of datasets focused on the **lungs, respiratory system, and associated diseases**. It includes data related to chronic respiratory conditions such as **asthma, chronic bronchitis, and COPD**, as well as infectious diseases like **tuberculosis, pneumonia, influenza, and whooping cough**. In addition, it provides datasets on risk factors and interventions, including smoking habits and nicotine replacement therapies, which are critical in understanding the epidemiology and prevention of respiratory illnesses.

## Dataset Suffixes

Each dataset in the `PulmoDataSets` package uses a `suffix` to denote the type of R object:

- `_df`: data frame

- `_dt`: data table

- `_tbl_df`: tibble

- `_ts`: time series

Below are selected example datasets included in the `PulmoDataSets` package:

- `bronchitis_Cardiff_df`: Chronic Bronchitis in Cardiff Men.

- `smoking_UK_tbl_df`: UK Smoking Habits.

- `nicotine_gum_df`: Nicotine Gum and Smoking Cessation.


## Data Visualization with PulmoDataSets Data

###  Chronic Bronchitis in Cardiff Men

```{r bronchitis-Cardiff-plot, fig.width=6, fig.height=4.5, out.width="90%"}

# Summary with .groups = "drop" to avoid the message (stored but not printed)
summary_stats <- bronchitis_Cardiff_df %>%
  group_by(r, rfac) %>%
  summarise(
    mean_cig = mean(cig, na.rm = TRUE),
    mean_poll = mean(poll, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )

# Plot only
ggplot(bronchitis_Cardiff_df, aes(x = cig, y = poll, color = factor(r))) +
  geom_point() +
  labs(
    title = "Cigarette Consumption vs Pollution",
    x = "Cigarette Consumption",
    y = "Pollution Level",
    color = "Bronchitis"
  ) +
  theme_minimal()

```


### UK Smoking Habits

```{r UK-smoking-plot, fig.width=6, fig.height=4.5, out.width="90%"}

smoking_summary <- smoking_UK_tbl_df %>%
  group_by(gender, smoke) %>%
  summarise(avg_amt_weekends = mean(amt_weekends, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(avg_amt_weekends))

ggplot(smoking_summary, aes(x = gender, y = avg_amt_weekends, fill = smoke)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Weekend Smoking by Gender and Smoking Status",
    x = "Gender",
    y = "Average Cigarettes on Weekends",
    fill = "Smoke Status"
  ) +
  theme_minimal()

```


### Nicotine Gum and Smoking Cessation

```{r nicotine-gum-plot, fig.width=6, fig.height=4.5, out.width="90%"}


# Step 1: Calculate mean success rates (no extra packages)
nicotine_summary <- nicotine_gum_df %>%
  summarize(
    treatment = sum(qt) / sum(tt),  # Overall success rate (treatment)
    control = sum(qc) / sum(tc)      # Overall success rate (control)
  )

# Step 2: Plot (manually reshape data without tidyr)
ggplot(data.frame(
  group = c("Treatment", "Control"),
  success_rate = c(nicotine_summary$treatment, nicotine_summary$control)
), aes(group, success_rate, fill = group)) +
  geom_col(width = 0.5) +
  labs(
    title = "Nicotine Gum vs. Control (Overall Success Rate)",
    y = "Success Rate",
    x = ""
  ) +
  scale_fill_manual(values = c("Treatment" = "#1f77b4", "Control" = "#d62728")) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Conclusion

`PulmoDataSets` package delivers ready-to-use respiratory datasets **(COPD, asthma, TB, pneumonia, etc.)** for clinical and epidemiological research. The package simplifies data access for modeling, teaching, and public health studies.

For detailed information and full documentation of each dataset, please refer to the reference manual and help files included within the package.

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
